Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-25 src/5.Example.tex
Improvements to Chapter on Example and Backend

In order to provide an impression of the looks and feel of Kokkos code this section discusses a small but complete example of a data parallel Kokkos application shown in Figure~\ref{fig:KokkosExample} which allocates data, potentially initalizes it on the host and then performs a matrix-vector product.
Similar to other programming models such as MPI, Kokkos must be initialized and finalized. 
This serves the purpose of setting up the underlying runtime and acquiring resouces.
The fundamental data structure of Kokkos programs is the ~\emph{Kokkos::View}.
These views represent multi-dimensional arrays of rank one to eight.
A general philosophy of Kokkos is to have defaults for most properties, including data layouts, memory spaces and execution spaces.
Consequently the simple example does not explicitly set which memory space the data lives in.
Kokkos ensures that the defaults for parallel execution and data allocations match though.

The example then creates copies of the views accessible by the host process via the ~\emph{create\_mirror\_view} function. 
If the original data was already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation.
In real applications this host process accessible allocation is often necessary for I/O purposes for example to initialize the data with input read from a file.
To get the data to the memory space accessible by the parallel execution resources the function ~\emph{deep\_copy} is then invoked.
If two ~\emph{View}s are aliases of each other this operation is a no-op.

Now the parallel pattern can be invoked.
Since the matrix-vector product is an example of hierarchical work (for each row a dot product must be performed), the ~\emph{Kokkos::TeamPolicy} is used.
For each row of M a team of threads is launched, where the size of team is left for the Kokkos runtime to choose.
When using a ~\emph{TeamPolicy} the operator of the lambda does not simply get an index, but rather a handle to the team of threads. 
This handle provides the team id, and is subsequently handed to the nested parallel execution policies. 
The nested reduction requires the operator of the nested lambda to take a reference to the thread local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index.
It then writes the result into the correct position of the output vector.

After the parallel pattern the main code calls a fence to ensure that the execution of the parallel code is finished, before copying the results of y back to the host and potentially putting them out through an I/O operation.

To learn more about Kokkos's capabilties and features we recommend going through the extensive Kokkos tutorial, with multiple days worth of lectures and exercises, and read the Kokkos Programming Guide as well as the Kokkos API documentation - which provide several hundred pages of text-book like material and API reference.
The Tutorial is intended for computational scientists with minimal or even no prior knowlegde of parallel programming.
It introduces concepts of parallel programming using Kokkos step by step, through a series of lectures and hands-on exercises which build on each other and gradually introduce more and more Kokkos constructs.
The Programming Guide provides more in-depth information to the concepts introduced in the Tutorial, while the API reference provides straight forward documentation of the syntax and semantics of Kokkos's library interface.

@@

The following section gives a short impression of Kokkos maps generic parallel programming patterns to the hardware specific programming models.
As a template metaprogramming library, Kokkos makes heavy use of partial specialization. 
In particular the parallel pattern functions are mapped to partial specializations of classes for those patterns for each execution policy type and backend type.
The library specification contains concepts for execution spaces, which define the generic interface of those classes.
To add a new backend, each of those concepts must be implemented.

A simplified version of this implementation strategy is provided in Figure\ref{fig:KokkosExampleOMPBackEnd}.
The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function.
That function is then responsible to implement the Kokkos parallel pattern, for example via OpenMP pragmas in the case of the OpenMP execution space.
\begin{figure}
\begin{small}
\begin{Verbatim}[frame=leftline]
template <class FunctorType>
class ParallelFor<FunctorType, RangePolicy<OpenMP> > {
  const FunctorType functor;
  const RangePolicy<OpenMP> policy;

@@

\end{Verbatim}
\end{small}
\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programing model. It shows a particular implementation of that generic pattern as a templated class to the OpenMP execution space. Also it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
\label{fig:KokkosExampleOMPBackEnd}
\end{figure}

>>>>

In this section we provide an example of an application written in the Kokkos programming model. The code in Figure~\ref{fig:KokkosExample} shows a parallel matrix-vector multiplication implemented as two nested loops. The outer loop, that is the outer parallel_for, iterates over the rows of the input matrix and of the input vector while the inner loop iterates over elements in each row. Since the inner loop potentially executes in parallel, the update operation of the aggregating variable~\emph{y_tmp} must be protected against data races. Using the parallel_reduction pattern, the programming model provides a local variable to that loop iteration. Once the computation finishes, the final result of the reduction operation is stored in the output vector~\emph{y} at the row index~\emph{e}.  
As a first step, a Kokkos application must call ~\emph{Kokkos::initialize}. Initialization and finalization are required and serve the purpose of setting up the underlying runtime. 
Further, the example code shows the use of the ~\emph{Kokkos::View} abstraction.

Views are abstractions that represent data. Naturally data can reside in host memory where it is accessible to the host execution space or on device memory where it is accessible to the device execution space. In this example, copies of the views are created in order to make them accessible by the host process. This is accomplished by calling the ~\emph{create\_mirror\_view} function.

If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation. In real applications host-accessible allocation is necessary in order to allow I/O operations during fileystem access. To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked. In case two view objects are aliased, this operation results in a~\emph{no-op}.

%A general philosophy of Kokkos is to have defaults for most properties, including data layouts, memory spaces and execution spaces.
Consequently the simple example does not explicitly set which memory space the data lives in.
Kokkos ensures that the defaults for parallel execution and data allocations match though.

Taking a closer look at the loop step shows the hierarchical expression of concurrency. In case of the matrix-vector multiplication,This is accomplished by using the ~\emph{Kokkos::TeamPolicy}. For each row of~\emph{M} a team of threads is launched, where the size of the team is not further specified. When using a ~\emph{TeamPolicy} the operator of the lambda expression does not receive an index, but rather a handle to the team of threads. 
This handle provides the team identifier \emph{e}, and is subsequently handed to the nested parallel execution policies. The nested reduction requires the operator of the nested lambda to take a reference to the thread-local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index. It then writes the result to the correct position of the output vector. 
The parallel loops are follows by a call to \emph{Kokkos::fence}. This ensures that the execution of the parallel code is completed before copying the results back to the host memory space. This is implemented through the~\emph{Kokkos::deep_copy} call. Lastly, the Kokkos application calls~\emph{Kokkos::finalize}.

While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources. Resources include example applications for individual Kokkos features, tutorials with multiple days worth of lectures and exercises, a programming guide and an API documentation. Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}. The tutorial is intended for students with minimal or even no prior knowledge of parallel programming. It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises which build up on each other and gradually introduce new concepts. The programming guide is available on-line~\cite{KOKKOS_WIKI}.

@@

In this section we peak inside the Kokkos library and show how the a generic parallel for loop is mapped to the OpenMP back-end programming model. As a template metaprogramming library, Kokkos makes use of partial specialization. In particular the~\emph{parallel_for} or ~\emph{parallel_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes. Such classes correspond to particular back-end where for each back-end, concepts with partial specialization must be provided. Today, Kokkos supports the CUDA and OpenMP back-ends. For each new back-end, partially implemented classes that implement necessary concepts must be provided.

A simplified version of a partially specialized class for a parallel for loop implementation is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}. The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function. That function is then responsible to implement the Kokkos parallel pattern. In this back-end, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled. 
\begin{figure}
\begin{small}
\begin{Verbatim}[frame=leftline]
template <class FunctorType>
class ParallelFor<FunctorType, RangePolicy<OpenMP> > {
  const FunctorType functor;
  const RangePolicy<OpenMP> policy;

@@

\end{Verbatim}
\end{small}
\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programing model. It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space. Also it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
\label{fig:KokkosExampleOMPBackEnd}
\end{figure}
